---
title: "Complementary diagnostic plots and predictive plots of linear mixed models from Lme4 "
author: "Yian Guo, Yuting Yang, Xiaoman Zhai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# 1. resid.plot

## Title: 
resid.plot  
Plotting pearson residuals for linear mixed effect models with outliers denotation

## Description:  
This function plots the pearson type residuals against fitted values from a linear mixed effect model fitted by lmer function, with optional denotation of possible outliers.

## Usage:
resid.plot(object, denote=TRUE)

## Arguments:
object: a lmer model
denote: gives the index with top 5% type pearson residuals, default = TRUE

## Example:
```{r, include=FALSE}
library(lme4)
library(lattice)
library(ggplot2)
library(tidyr)
library(HLMdiag)

resid.plot <- function(mod, denote = TRUE) {
    red <- data.frame(residual = resid(mod, type = "pearson"), fitted = fitted(mod))
    cut <- quantile(abs(red$residual), prob = 0.975)
    if (denote == TRUE) {
        print(ggplot(red, aes(fitted, residual)) + geom_point() + geom_text(aes(label = ifelse(abs(residual) >= 
            cut, rownames(red), "")), hjust = -0.3, vjust = -0.3, size = 3) + theme_bw() + geom_hline(yintercept = 0, 
            color = "blue"))
    } else {
        print(ggplot(red, aes(fitted, residual)) + geom_point() + theme_bw() + geom_hline(yintercept = 0, 
            color = "blue"))
    }
}
```
```{r}
mod<-lmer(Reaction ~ Days + (Days|Subject), data=sleepstudy)
resid.plot(mod, denote=TRUE)
```

# 2. lev.plot

## Title: lev.plot 
Plotting leverages against pearson residuals for linear mixed effect models  

## Description: 
This function plots the leverages (calculated by HLMdiag ::leverage) against pearson residuals from a linear mixed effect model fitted by lmer function.

## Usage:
lev.plot(object, type="overall")

## Arguments:
object: a lmer model
type: "overall" plots the overall leverage; "fixef" plots the leverage corresponding to the fixed effects; "ranef" plots the leverage corresponding to the random effects. Default is "overall".

## Example:
```{r, include=FALSE}
lev.plot <- function(mod, type = "overall") {
    lev <- leverage(mod, level = 1)
    colnames(lev) <- c("overall leverage", "fixed effect leverage", "random effect leverage")
    if (type == "overall") {
        ggplot(data.frame(lev = lev[, 1], pearson = residuals(mod, type = "pearson")), aes(x = lev, 
            y = pearson)) + geom_point() + theme_bw() + ggtitle(colnames(lev)[1])
    } else if (type == "fixef") {
        ggplot(data.frame(lev = lev[, 2], pearson = residuals(mod, type = "pearson")), aes(x = lev, 
            y = pearson)) + geom_point() + theme_bw() + ggtitle(colnames(lev)[2])
    } else if (type == "ranef") {
        ggplot(data.frame(lev = lev[, 3], pearson = residuals(mod, type = "pearson")), aes(x = lev, 
            y = pearson)) + geom_point() + theme_bw() + ggtitle(colnames(lev)[3])
    }
}
```
```{r}
mod<-lmer(Reaction ~ Days + (Days|Subject), data=sleepstudy)
lev.plot(mod, type="fixef")
```

# 3. cooks.plot

## Title: cooks.plot  
Plotting cook's distance for linear mixed effect models.

## Description: 
This function plots the cook's distance (calculated by HLMdiag ::cooks.distance) a linear mixed effect model fitted by lmer function.

## Usage:
cooks.plot(object, data, cut)

## Arguments:
object: a lmer model
data: the dataset used in the object
cut: cutoff value, unless specified, equals 4/n where n equals the observation number in the dataset.

## Example:
```{r, include=FALSE}
cooks.plot <- function(mod, data, cut) {
    cook <- cooks.distance(mod)
    cook.df <- data.frame(Index = 1:dim(data)[1], cooks.distance = cook[1:dim(data)[1]])
    if (missing(cut)) {
        ggplot(cook.df, aes(Index, cooks.distance)) + geom_point() + geom_text(aes(label = ifelse(cooks.distance >= 
            4/dim(data)[1], rownames(cook.df), "")), hjust = -0.3, vjust = -0.3, size = 3) + theme_bw() + 
            geom_hline(yintercept = 4/dim(data)[1], color = "blue")
    } else {
        ggplot(cook.df, aes(Index, cooks.distance)) + geom_point() + geom_text(aes(label = ifelse(cooks.distance >= 
            cut, rownames(cook.df), "")), hjust = -0.3, vjust = -0.3, size = 3) + theme_bw() + geom_hline(yintercept = cut, 
            color = "blue")
    }
}
```
```{r}
mod<-lmer(Reaction ~ Days + (Days|Subject), data=sleepstudy)
cooks.plot(mod, data=sleepstudy)
```
# 4. pred.plot

## Title:
pred.plot generate prediction and observation plots 

## Description: 
This function plots predicted values and observed response values against time variable associated with repeated measures from a linear mixed effect model for visual comparison of model fit.

## Usage:
pred.plot(mod, data, subject, time, response)

# Arguments:
object: a lmer model
data: the dataset used in the object
subject: name of the independent sampling unit (subject, person, ID.)
time: name of the time variable associated with repeated measures (time, days, visit)
response: name of the response variable

## Example:
```{r,include=FALSE}
pred.plot <- function(mod, data, subject, time, response) {
    pred <- predict(mod)
    actual <- data[, response]
    sub <- data[, subject]
    tim <- data[, time]
    dat <- within(data, {
        Predicted <- pred
        Observed <- actual
        Subject <- sub
        Time <- tim
    })
    dat2 <- gather(dat, key = Class, value = Value, Observed, Predicted)
    ggplot(dat2) + aes(Time, Value, colour = Subject) + geom_line() + facet_wrap(~Class) + theme_bw()
}
```
```{r}
mod<-lmer(Reaction ~ Days + (Days|Subject), data=sleepstudy)
pred.plot(mod, sleepstudy, subject="Subject", time="Days", response="Reaction")
```

